---
title: "Project"
author: "Daniel Vasilevko"
date: "2026-02-20"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r}
library(jsonlite)
library(tibble)
library(dplyr)
library(tidytext)
library(ggplot2)
library(broom)
library(tidyr)
library(lexicon)
library(stringr)
```

```{r}
data <- stream_in(file("News.json"))
```

```{r}
news_tbl <- as_tibble(data)
head(news_tbl)
unique(news_tbl$category)

news_tbl %>%
  filter(category == "TRAVEL") %>%
  nrow()

dim(news_tbl)

head(news_tbl)
```

# Pie Chart Template

```{r}
world_tbl <- news_tbl %>%
  filter(category == "WORLD NEWS")

world_words <- world_tbl %>%
  select(link, headline) %>%
  unnest_tokens(word, headline)

bing <- get_sentiments("bing")

headline_sentiment <- world_words %>%
  inner_join(bing, by = "word")

headline_scores <- headline_sentiment %>%
  count(link, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(score = positive - negative)

headline_scores <- headline_scores %>%
  mutate(label = case_when(
    score > 0 ~ "Positive",
    score < 0 ~ "Negative",
    TRUE ~ "Neutral"
  ))

pie_data <- headline_scores %>%
  count(label)

pie_data <- pie_data %>%
  mutate(
    percent = round(100 * n / sum(n), 1),
    label_text = paste0(label, " (", percent, "%)")
  )

pie_data <- pie_data %>%
  mutate(label = factor(label,
                        levels = c("Negative", "Neutral", "Positive")))

ggplot(pie_data, aes(x = "", y = n, fill = label)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label_text),
            position = position_stack(vjust = 0.5),
            size = 4,
            fontface = "bold") +
  scale_fill_manual(values = c(
    "Negative" = "#E63946",  
    "Neutral"  = "#F1FAEE", 
    "Positive" = "#1D9BF0"   
  )) +
  theme_void() +
  labs(
    title = "Sentiment Distribution of World News Headlines",
    fill = "Sentiment"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "bottom", text = element_text(family = "Helvetica")
  )
```

```{r, echo= FALSE}
world_tbl <- news_tbl %>%
  filter(category == "U.S. NEWS")

world_words <- world_tbl %>%
  select(link, headline) %>%
  unnest_tokens(word, headline)

bing <- get_sentiments("bing")

headline_sentiment <- world_words %>%
  inner_join(bing, by = "word")

library(tidyr)

headline_scores <- headline_sentiment %>%
  count(link, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(score = positive - negative)

headline_scores <- headline_scores %>%
  mutate(label = case_when(
    score > 0 ~ "Positive",
    score < 0 ~ "Negative",
    TRUE ~ "Neutral"
  ))

pie_data <- headline_scores %>%
  count(label)

pie_data <- pie_data %>%
  mutate(
    percent = round(100 * n / sum(n), 1),
    label_text = paste0(label, " (", percent, "%)")
  )

pie_data <- pie_data %>%
  mutate(label = factor(label,
                        levels = c("Negative", "Neutral", "Positive")))

ggplot(pie_data, aes(x = "", y = n, fill = label)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label_text),
            position = position_stack(vjust = 0.5),
            size = 4,
            fontface = "bold") +
  scale_fill_manual(values = c(
    "Negative" = "#E63946",  
    "Neutral"  = "#F1FAEE", 
    "Positive" = "#1D9BF0"   
  )) +
  theme_void() +
  labs(
    title = "Sentiment Distribution of U.S. News Headlines",
    fill = "Sentiment"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "bottom", text = element_text(family = "Helvetica")
  )
```


```{r,echo=FALSE}
world_tbl <- news_tbl %>%
  filter(category == "ENTERTAINMENT")

world_words <- world_tbl %>%
  select(link, headline) %>%
  unnest_tokens(word, headline)

bing <- get_sentiments("bing")

headline_sentiment <- world_words %>%
  inner_join(bing, by = "word")

library(tidyr)

headline_scores <- headline_sentiment %>%
  count(link, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(score = positive - negative)

headline_scores <- headline_scores %>%
  mutate(label = case_when(
    score > 0 ~ "Positive",
    score < 0 ~ "Negative",
    TRUE ~ "Neutral"
  ))

pie_data <- headline_scores %>%
  count(label)

pie_data <- pie_data %>%
  mutate(
    percent = round(100 * n / sum(n), 1),
    label_text = paste0(label, " (", percent, "%)")
  )

pie_data <- pie_data %>%
  mutate(label = factor(label,
                        levels = c("Negative", "Neutral", "Positive")))

ggplot(pie_data, aes(x = "", y = n, fill = label)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label_text),
            position = position_stack(vjust = 0.5),
            size = 4,
            fontface = "bold") +
  scale_fill_manual(values = c(
    "Negative" = "#E63946",  
    "Neutral"  = "#F1FAEE", 
    "Positive" = "#1D9BF0"   
  )) +
  theme_void() +
  labs(
    title = "Sentiment Distribution of Entertainment Headlines",
    fill = "Sentiment"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "bottom", text = element_text(family = "Helvetica")
  )
```

```{r,echo=FALSE}
world_tbl <- news_tbl %>%
  filter(category == "TRAVEL")

world_words <- world_tbl %>%
  select(link, headline) %>%
  unnest_tokens(word, headline)

bing <- get_sentiments("bing")

headline_sentiment <- world_words %>%
  inner_join(bing, by = "word")

library(tidyr)

headline_scores <- headline_sentiment %>%
  count(link, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(score = positive - negative)

headline_scores <- headline_scores %>%
  mutate(label = case_when(
    score > 0 ~ "Positive",
    score < 0 ~ "Negative",
    TRUE ~ "Neutral"
  ))

pie_data <- headline_scores %>%
  count(label)

pie_data <- pie_data %>%
  mutate(
    percent = round(100 * n / sum(n), 1),
    label_text = paste0(label, " (", percent, "%)")
  )

pie_data <- pie_data %>%
  mutate(label = factor(label,
                        levels = c("Negative", "Neutral", "Positive")))

ggplot(pie_data, aes(x = "", y = n, fill = label)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label_text),
            position = position_stack(vjust = 0.5),
            size = 4,
            fontface = "bold") +
  scale_fill_manual(values = c(
    "Negative" = "#E63946",  
    "Neutral"  = "#F1FAEE", 
    "Positive" = "#1D9BF0"   
  )) +
  theme_void() +
  labs(
    title = "Sentiment Distribution of Travel Headlines",
    fill = "Sentiment"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "bottom", text = element_text(family = "Helvetica")
  )
```

# Headline Pos/Neg Scoring Bing

```{r}
relevant_tbl <- news_tbl %>%
  filter(category %in% c("WORLD NEWS", "U.S. NEWS", "TRAVEL", "ENTERTAINMENT"))

relevant_words <- relevant_tbl %>%
  select(link, headline, category) %>%
  unnest_tokens(word, headline)

bing <- get_sentiments("bing")

headline_sentiment <- relevant_words %>%
  inner_join(bing, by = "word")

headline_scores <- headline_sentiment %>%
  count(link, category, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(score = positive - negative) %>%
  mutate(label = case_when(
    score > 0 ~ "Positive",
    score < 0 ~ "Negative",
    TRUE ~ "Neutral"
  ))

headline_scores
```

# Headline NRC Emotion Words Linear Reg

```{r}
nrc <- get_sentiments("nrc")

headline_words <- relevant_tbl %>%
  select(link, headline, category) %>%
  unnest_tokens(word, headline)

headline_nrc <- headline_words %>%
  inner_join(nrc, by = "word")

headline_emotions <- headline_nrc %>%
  count(link, category, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  )

headline_emotions <- headline_emotions %>%
  mutate(new_category = case_when(
    category %in% c("WORLD NEWS", "U.S. NEWS") ~ "News",
    category %in% c("ENTERTAINMENT", "TRAVEL") ~ "General",
    TRUE ~ category  
  ))

head(headline_emotions)

model_anger <- lm(anger ~ new_category, data = headline_emotions)
summary(model_anger)
tidy(model_anger)

model_disgust <- lm(disgust ~ new_category, data = headline_emotions)
summary(model_disgust)
tidy(model_disgust)

model_fear <- lm(fear ~ new_category, data = headline_emotions)
summary(model_fear)
tidy(model_fear)

model_sadness <- lm(sadness ~ new_category, data = headline_emotions)
summary(model_sadness)
tidy(model_sadness)

model_joy <- lm(joy ~ new_category, data = headline_emotions)
summary(model_joy)
tidy(model_joy)

model_anticipation <- lm(anticipation ~ new_category, data = headline_emotions)
summary(model_anticipation)
tidy(model_anticipation)

model_surprise <- lm(surprise ~ new_category, data = headline_emotions)
summary(model_surprise)
tidy(model_surprise)

model_trust <- lm(trust ~ new_category, data = headline_emotions)
summary(model_trust)
tidy(model_trust)


```

```{r}
emotion_data_all <- headline_emotions %>%
  mutate(new_category = case_when(
    category %in% c("WORLD NEWS", "U.S. NEWS") ~ "Hard News",
    category %in% c("ENTERTAINMENT", "TRAVEL") ~ "Lifestyle News"
  )) %>%
    pivot_longer(
    cols = -c(link, category, new_category),
    names_to = "emotion",
    values_to = "count"
  ) %>%
  group_by(new_category, emotion) %>%
  summarise(avg_count = mean(count), .groups = "drop")

ggplot(emotion_data_all, aes(x = emotion, y = avg_count, fill = new_category)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("Hard News" = "#E63946", "Lifestyle News" = "#1D9BF0")) + 
  labs(
    title = "Mean Emotive Words per Headline by Category",
    x = "Emotion",
    y = "Average Count of Words",
    fill = "Category"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Helvetica", color = "white"),
    plot.title = element_text(face = "bold", hjust = 0.5, color = "white"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    panel.background = element_rect(fill = "#1d3159", color = NA),
    plot.background = element_rect(fill = "#1d3159", color = NA),
    legend.background = element_rect(fill = "#1d3159", color = NA),
    legend.key = element_rect(fill = "#1d3159", color = NA)
  )
```

# Hard vs Soft News Pos/Neg Linear Regression

```{r}
headline_scores <- headline_scores %>%
  mutate(new_category = case_when(
    category %in% c("WORLD NEWS", "U.S. NEWS") ~ "Hard News",
    category %in% c("ENTERTAINMENT", "TRAVEL") ~ "General News",
    TRUE ~ category  
  ))

table(headline_scores$new_category)

model <- lm(score ~ new_category, data = headline_scores)

summary(model)

tidy(model)
```

# Headline and Summary Pos/Neg Scoring

```{r}
relevant_tbl <- news_tbl %>%
  filter(category %in% c("WORLD NEWS", "U.S. NEWS", "TRAVEL", "ENTERTAINMENT"))

bing <- get_sentiments("bing")

headline_scores <- relevant_tbl %>%
  select(link, headline, category) %>%
  unnest_tokens(word, headline) %>%
  inner_join(bing, by = "word") %>%
  count(link, category, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(headline_score = positive - negative) %>%
  select(-positive, -negative)  

summary_scores <- relevant_tbl %>%
  select(link, short_description) %>%
  unnest_tokens(word, short_description) %>%
  inner_join(bing, by = "word") %>%
  count(link, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(article_score = positive - negative) %>%
  select(link, article_score)

combined_scores <- headline_scores %>%
  left_join(summary_scores, by = "link")

combined_scores <- combined_scores %>%
  mutate(article_score = replace_na(article_score, 0))

combined_scores
```

# Headline Summary Pos/Neg T-Test

```{r}
hard_news <- combined_scores %>%
  filter(category %in% c("WORLD NEWS", "U.S. NEWS"))

hard_news <- hard_news %>%
  mutate(diff = headline_score- article_score )

t.test(hard_news$headline_score,
       hard_news$article_score,
       paired = TRUE)

general_news <- combined_scores %>%
  filter(category %in% c("TRAVEL", "ENTERTAINMENT"))

t.test(general_news$headline_score,
       general_news$article_score,
       paired = TRUE)

```

# Headline Summary Pos/Neg Linear Regression

```{r}
combined_scores <- combined_scores %>%
  mutate(
    group = ifelse(category %in% c("WORLD NEWS", "U.S. NEWS"),
                   "Hard News",
                   "General News"),
    diff = headline_score - article_score
  )

summary(lm(diff ~ group, data = combined_scores))

group_means <- combined_scores %>%
  group_by(group) %>%
  summarise(
    mean_diff = mean(diff),
    mean_headline = mean(headline_score),
    mean_summary = mean(article_score)
  )

ggplot(combined_scores, aes(x = group, y = diff)) +
  geom_jitter(width = 0.2, alpha = 0.2, color = "#1D9BF0") + 
  geom_point(data = group_means, aes(x = group, y = mean_diff), color = "#E63946", size = 4) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "white") +
  labs(
    title = "Headline vs Summary Sentiment Difference by Category",
    x = "Category",
    y = "Headline - Summary Sentiment Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Helvetica", color = "white"),
    plot.title = element_text(face = "bold", hjust = 0.5, color = "white"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    panel.background = element_rect(fill = "#1d3159", color = NA),
    plot.background = element_rect(fill = "#1d3159", color = NA)
  )
```

# Headline Hard vs Soft News Intensifier Scoring

```{r}
additional_intensifiers <- c(
  "amazingly",
  "astonishingly",
  "astoundingly",
  "blatantly",
  "dramatically",
  "drastically",
  "enormous",
  "enormously",
  "exceedingly",
  "extraordinarily",
  "fantastically",
  "frightfully",
  "gigantic",
  "gigantically",
  "inordinately",
  "insanely",
  "intensely",
  "literally",
  "mind-blowingly",
  "notably",
  "outrageously",
  "phenomenal",
  "phenomenally",
  "profoundly",
  "radically",
  "remarkably",
  "ridiculously",
  "sharply",
  "significantly",
  "staggeringly",
  "substantially",
  "supremely",
  "tremendous",
  "tremendously",
  "unbelievably",
  "undeniably",
  "unusually",
  "utter",
  "utterly",
  "vast",
  "vastly",
  "wildly"
)

shifters <- hash_valence_shifters

custom_intensifiers <- unique(c(
  hash_valence_shifters %>% 
    filter(y == 2) %>% 
    pull(x),
  additional_intensifiers
))

custom_shifters <- tibble(
  x = additional_intensifiers,
  y = "2"
)

shifters <- bind_rows(shifters, custom_shifters) %>%
  distinct(x, .keep_all = TRUE)

```

```{r}
# https://www.rdocumentation.org/packages/lexicon/versions/1.2.1/topics/hash_valence_shifters
relevant_tbl <- news_tbl %>%
  filter(category %in% c("WORLD NEWS", "U.S. NEWS",
                         "TRAVEL", "ENTERTAINMENT"))

relevant_words <- relevant_tbl %>%
  select(link, headline, category) %>%
  unnest_tokens(word, headline)

headline_intensifiers <- relevant_words %>%
  inner_join(shifters, by = c("word" = "x")) %>%
  filter(y == 2)

headline_intensifier_counts <- headline_intensifiers %>%
  count(link, category) %>%
  rename(intensifiers = n)

headline_intensifier_counts <- relevant_tbl %>%
  select(link, category) %>%
  distinct() %>%
  left_join(headline_intensifier_counts,
            by = c("link", "category")) %>%
  mutate(intensifiers = ifelse(is.na(intensifiers), 0, intensifiers))

headline_intensifier_counts <- headline_intensifier_counts %>%
  mutate(new_category = case_when(
    category %in% c("WORLD NEWS", "U.S. NEWS") ~ "Hard News",
    category %in% c("ENTERTAINMENT", "TRAVEL") ~ "General News"
  ))

intense <- shifters[shifters$y ==2]

```

```{r}
headline_intensifier_counts <- headline_intensifier_counts %>%
  mutate(has_intensifier = ifelse(intensifiers > 0, 1, 0))

prop_data <- headline_intensifier_counts %>%
  group_by(new_category) %>%
  summarise(
    proportion = mean(has_intensifier)
  )

ggplot(prop_data,
       aes(x = new_category, y = proportion, fill = new_category)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(
    "Hard News" = "#1D9BF0",
    "General News" = "#E63946"
  )) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Proportion of Headlines Containing Intensifiers",
    x = "Category",
    y = "Percentage of Headlines"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Helvetica", color = "white"),
    plot.title = element_text(face = "bold", hjust = 0.5, color = "white"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),

    legend.position = "none",

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    panel.background = element_rect(fill = "#1d3159", color = NA),
    plot.background = element_rect(fill = "#1d3159", color = NA)
  )
```

# Headline Hard vs Soft News Intensifier Linear Regression

```{r}
model <- lm(intensifiers ~ new_category,
            data = headline_intensifier_counts)

summary(model)
```

# Headline Hard vs Soft News Downtoner Scoring

```{r}
headline_downtoner <- relevant_words %>%
  inner_join(shifters, by = c("word" = "x")) %>%
  filter(y == 3)

headline_downtoner_counts <- headline_downtoner %>%
  count(link, category) %>%
  rename(downtoner = n)

headline_downtoner_counts <- relevant_tbl %>%
  select(link, category) %>%
  distinct() %>%
  left_join(headline_downtoner_counts,
            by = c("link", "category")) %>%
  mutate(downtoner = ifelse(is.na(downtoner), 0, downtoner))

headline_downtoner_counts <- headline_downtoner_counts %>%
  mutate(new_category = case_when(
    category %in% c("WORLD NEWS", "U.S. NEWS") ~ "Hard News",
    category %in% c("ENTERTAINMENT", "TRAVEL") ~ "General News"
  ))

```


```{r}
headline_downtoner_counts <- headline_downtoner_counts %>%
  mutate(has_downtoner = ifelse(downtoner > 0, 1, 0))

prop_downtoner <- headline_downtoner_counts %>%
  group_by(new_category) %>%
  summarise(
    proportion = mean(has_downtoner)
  )

ggplot(prop_downtoner,
       aes(x = new_category, y = proportion, fill = new_category)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(
    "Hard News" = "#1D9BF0",
    "General News" = "#E63946"
  )) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Proportion of Headlines Containing Downtoners",
    x = "Category",
    y = "Percentage of Headlines"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Helvetica", color = "white"),
    plot.title = element_text(face = "bold", hjust = 0.5, color = "white"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white"),

    legend.position = "none",

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    panel.background = element_rect(fill = "#1d3159", color = NA),
    plot.background = element_rect(fill = "#1d3159", color = NA)
  )
```


# Headline Hard vs Soft News Downtoner Linear Regression


```{r}
model <- lm(downtoner ~ new_category,
            data = headline_downtoner_counts)

summary(model)
```

# Word Clouds

```{r}
library(dplyr)
library(tidytext)
library(wordcloud)
library(RColorBrewer)
library(stringr)

headline_words <- relevant_tbl %>%
  select(link, headline, category) %>%
  unnest_tokens(word, headline) %>%
  mutate(new_category = case_when(
    category %in% c("WORLD NEWS", "U.S. NEWS") ~ "News",
    category %in% c("ENTERTAINMENT", "TRAVEL") ~ "General"
  ))

data("stop_words")
headline_words <- headline_words %>%
  anti_join(stop_words, by = "word") %>%
  filter(!str_detect(word, "^[0-9]+$"))  # remove words that are only numbers


set.seed(123)

par(bg = "#1d3159")

news_words <- headline_words %>% filter(new_category == "News")

news_words %>%
  count(word, sort = TRUE) %>%
  with(wordcloud(
    words = word,
    freq = n,
    max.words = 100,
    colors = brewer.pal(8, "RdYlBu"),
    random.order = FALSE
  ))

set.seed(123)

par(bg = "#1d3159")


general_words <- headline_words %>% filter(new_category == "General")

general_words %>%
  count(word, sort = TRUE) %>%
  with(wordcloud(
    words = word,
    freq = n,
    max.words = 100,
    scale = c(3, 0.35),
    colors = brewer.pal(8, "RdYlBu"),
    random.order = FALSE
  ))
```





